---
title: "Harmonic Regression"
author: "Austin Stephen"
date: "4/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
```

```{r include = FALSE}
# get the data
data <- read_csv("../data/SMP500_post_1900.csv")
```

## TLDR

* I implemented harmonic regression more or less by hand.

* It did not work very well. Probably because I don't know how to find a good 
period just looking at the data.

### Formulation:
$X_t = s_t + Y$ where $s_t$ is a periodic function and $Y$ is the noise term

$s_t = a_0 + \Sigma_{j=1}^k(a_j cos(\lambda_jt)+b_jsin(\lambda_j t))$

$a_1,a_2,...a_n$ and $b_1, b_2, ..,b_n$ are unknown parameters. $\lambda_1,..,\lambda_k$ are fixed frequencies as integer multiples of $2\pi / d$ where d is the period.


A harmonic regression model allows for noise but no trend component.
As a result, it is trained on the residuals of the multiple linear regression model used to subtract out the trend in the data. I fit the $a_j$ and $b_j$ components using the 
least squares method with the built in function in R.

The number of terms $k$ (indexing the summation) are refered to as the k-integer valued fourier components. I build a model with 1 fourier component and another with 2. Each fourier component must be 
a fixed integer multiple of $f_1$ which denotes the first component.

All of the math comes from Introduction to Time Series and Forecasting by Brockwell and Davis. Chapter 1.3 "Some simple models".


### One fourier component  

$s_t = a_0 + (a_1 cos(\lambda_jt)+b_1sin(\lambda_j t))$

```{r, fig.height=3.5}
## one fourier component
# Difficult to infer lambda based on our data
pi = 3.14159265
period = 4
lambdaTerm <- 2* pi / period

data <- data %>% mutate(aTerm = sin(lambdaTerm * dateNumeric),
                        bTerm = cos(lambdaTerm * dateNumeric))

model <- lm(residualsCube ~ aTerm + bTerm, data = data)
#summary(model)

## plot curve add the curve
data %>% ggplot(aes(x=dateNumeric,y=residualsCube))+
  geom_point()+
  theme_classic()
```

### Two Fourier Components   

$s_t = a_0 + (a_1 cos(\lambda_jt)+b_1sin(\lambda_j t)) + (a_2 cos(\lambda_jt)+b_2sin(\lambda_j t))$

```{r fig.height=3.5}
## two fourier components
# Difficult to infer lambda based on out data
period = 4

lambdaTerm1 <- 2* pi / period

## fixed integer multiple
lambdaTerm2 <- 2* pi / (period* 2)


data <- data %>% mutate(a1Term = sin(lambdaTerm1 * dateNumeric),
                        a2Term = sin(lambdaTerm2 * dateNumeric),
                        b1Term = cos(lambdaTerm1 * dateNumeric),
                        b2Term = cos(lambdaTerm2 * dateNumeric)
                  )

model2 <- lm(residualsCube ~ aTerm + bTerm + a2Term + b2Term, data = data)
#summary(model)

## plot curve -> add the curve 
data %>% ggplot(aes(x=dateNumeric,y=residualsCube))+
  geom_point()+
  theme_classic()
  
```

## Same thing but with data from 1950 on
```{r include = FALSE}
data_1950 <- read_csv("../data/SMP500_post_1950.csv")
```

```{r fig.height =3.5}
data_1950 <- data_1950 %>% mutate(aTerm = sin(lambdaTerm * dateNumeric),
                                  bTerm = cos(lambdaTerm * dateNumeric))

model <- lm(residualsCube ~ aTerm + bTerm, data = data_1950)
#summary(model)

## plot curve add the curve
data_1950 %>% ggplot(aes(x=dateNumeric,y=residualsCube))+
  geom_point()+
  theme_classic()

lambdaTerm1 <- 2* pi / period

## fixed integer multiple
lambdaTerm2 <- 2* pi / (period* 2)


data_1950 <- data_1950 %>% mutate(a1Term = sin(lambdaTerm1 * dateNumeric),
                        a2Term = sin(lambdaTerm2 * dateNumeric),
                        b1Term = cos(lambdaTerm * dateNumeric),
                        b2Term = cos(lambdaTerm2 * dateNumeric)
                  )

model2 <- lm(residualsCube ~ aTerm + bTerm + a2Term + b2Term, data = data_1950)
#summary(model)

## plot curve -> add the curve 
data_1950 %>% ggplot(aes(x=dateNumeric,y=residualsCube))+
  geom_point()+
  theme_classic()
```

### Train/test analysis

```{r}
## train on 1950-2005 -> predict 2005-current 16 years
train <- data_1950 %>% filter(dataNumeric < 48588)
test <- data_1950 %>% filter(dataNumeric >= 48588)

## model predictions 

# predictions %>% ggplot(aes(x=dateNumeric,y=pred))+
#   geom_point()+
#   theme_classic()

## convert back to original scale 
```

